---
title: "Final project"
output: html_document
date: "2022-12-11"
---

```{r}
cb_palette = c("#999999", "#E69F00", "#56B4E9", "#009E73",
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
packages = c('tidyverse', 
'ggplot2', 
'dplyr', 
'tidyr',
'plotly',
'RColorBrewer'
)
for (p in packages){
  library(p,character.only = T)
}
```


```{r}
df_happiness = read.csv('World_Happiness_index(2005 - 2021).csv', header = TRUE)
df_unemployment = read.csv('Unemployment_Rate.csv', header = TRUE)
df_continents = read.table("continents.txt",sep = '', header = TRUE)
```

Filtering the dataset
```{r}
df_happiness = df_happiness %>%
  filter(year >= 2010)

df_unemployment = df_unemployment %>%
  filter(Year >= 2010)


```

```{r}
df_unemployment = df_unemployment %>%
  rename(year = Year,
         Country = Entity,
         Unemployment_Rate = Unemployment..total....of.total.labor.force...modeled.ILO.estimate.,
         
         )

df_happiness = df_happiness %>%
  rename(
         Country = Country.name,
         GDP_percapita = Log.GDP.per.capita,
         Life_expextancy = Healthy.life.expectancy.at.birth,
         Corruption_rate = Perceptions.of.corruption
         )

df_continents = df_continents %>%
  rename(Country =  name)

```


```{r}
Final_df = merge(df_happiness,df_unemployment[,c("Country", "year", "Unemployment_Rate")],
                   by=c("Country", "year"))
```

```{r}
Final_df = merge(Final_df, df_continents[,c("Country", "continent")], by = c("Country"))
```


```{r}
Final_df = Final_df[,c("Country", "continent", "year", "Life.Ladder", "GDP_percapita", "Life_expextancy", "Unemployment_Rate")]

Final_df = na.omit(Final_df)
```

```{r}
length(unique(Final_df$Country))
```

## Map of average Happiness Index
```{r}
final_mean=
 Final_df %>%
  merge(df_continents[,c("Country", "continent")], by = c("Country")) %>%
  group_by(Country) %>%
  
  summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate), mean_gdp=mean(GDP_percapita),mean_life= mean(Life_expextancy))
```


```{r}
library(rworldmap)

ddf <- subset(final_mean[c("Country","mean_happiness")])
class(ddf)
```

```{r}
sPDF <- joinCountryData2Map( ddf
,joinCode = "NAME"
,nameJoinColumn = "Country")
```

```{r}
library(RColorBrewer)
colours=brewer.pal(4,"Spectral")
mapCountryData(sPDF, nameColumnToPlot="mean_happiness", catMethod="fixedWidth",colourPalette = colours)
```

```{r}
v_color <- viridis::viridis(
  n = nrow(Final_df)
)

Final_df$color <- v_color[Matrix::invPerm(
  p = order(
    x = Final_df$Life.Ladder
  )
)]
```


##Correlation of Happiness score and three dependent variables. 
```{r}
pairs(
  formula = Life.Ladder ~ GDP_percapita + Life_expextancy +
    Unemployment_Rate,
  data = Final_df,
  col = Final_df$color
,pch = 19
)
```




```{r}
final_mean_df= Final_df %>%
  group_by(Country) %>%
  summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate), mean_gdp=mean(GDP_percapita),mean_life= mean(Life_expextancy))%>%
   merge(df_continents[,c("Country", "continent")], by = c("Country"))

```

```{r}
final_mean_df=as.data.frame(final_mean_df)
```

```{r}
ggplot(final_mean_df , aes(x = continent, y = mean_happiness)) +
  geom_boxplot(aes(fill=continent)) + theme_bw() +
  theme(axis.text.x = element_text (angle = 90))+theme_dark()  +
  labs(y = "Average Happiness Score",
       x = "Continents",
       title = "Happiness with continents")
```




```{r}
final_mean2=
 Final_df %>%
  group_by(Country) %>%
  filter(year < 2019) %>%

  summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate), mean_gdp=mean(GDP_percapita),mean_life= mean(Life_expextancy))

```

```{r}
final_mean3=merge(final_mean2,df_continents, by="Country") 
```

```{r}
df <-final_mean3[order(final_mean3$mean_happiness),]
```

```{r}
final_mean3.top10 = tail(df,10)
```

```{r}
ggplot(final_mean3.top10, aes(x = reorder(Country,-mean_happiness), y = mean_happiness,fill=continent)) + geom_bar(stat="identity")+ theme(axis.text.x = element_text(angle = 45))+geom_col(colour = "black") + coord_cartesian(ylim=c(4,8))+scale_fill_manual(values = c("Asia"="#999999", "Europe"="#009E73", "Oceania"="#E69F00", "Americas"="#0072B2")) + labs(y = "Happiness Index",
       x = "Country",
       title = "Top 10 Happiest countries before COVID")
```

```{r}
final_mean4=
 Final_df %>%
  group_by(Country) %>%
  filter(year > 2019) %>%

  summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate), mean_gdp=mean(GDP_percapita),mean_life= mean(Life_expextancy))

```

```{r}
final_mean5=merge(final_mean4,df_continents, by="Country") 
```

```{r}
df2 <-final_mean5[order(final_mean5$mean_happiness),]
```

```{r}
final_mean5.top10 = tail(df2,10)
```

```{r}
ggplot(final_mean5.top10, aes(x = reorder(Country,-mean_happiness), y = mean_happiness,fill=continent)) + geom_bar(stat="identity")+ theme(axis.text.x = element_text(angle = 45))+geom_col(colour = "black")+coord_cartesian(ylim=c(4,8))+scale_fill_manual(values = c("Asia"="#999999", "Europe"="#009E73", "Oceania"="#E69F00", "Americas"="#0072B2"))+ labs(y = "Happiness Index",
       x = "Country",
       title = "Top 10 Happiest countries during COVID")
```

## Relationship among Unemployment Rate, GDP to Happiness Index

```{r}
Final_df %>%
  group_by(Country) %>%
  filter(year < 2019) %>%
  summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate), mean_gdp=mean(GDP_percapita),mean_life= mean(Life_expextancy))%>%
   merge(df_continents[,c("Country", "continent")], by = c("Country")) %>%
  ggplot(aes(x=mean_unemployment,y=mean_happiness,size=mean_gdp,col=continent))+geom_jitter(alpha=0.6)+ scale_color_manual(values = cb_palette)+geom_smooth(method="lm", span =0.8,se=FALSE) +
  labs(y = "Happiness Index",
       x = "Unemployment Rate",
       title = "Happines vs Unemployment with GDP as third variable")
```


## Time plot of average Happiness for each year

```{r}
Final_df$yearfactor = factor(Final_df$year)
Final_df %>%
  group_by(yearfactor) %>%
  summarise(mean_happiness = mean(Life.Ladder)) %>%
  ggplot(aes(x=yearfactor, y = mean_happiness))  + geom_point() + geom_line(aes(group=1)) 
```
## Time plot of how the Happiness varies over the years for each continent

```{r}
Final_df %>%
  group_by(continent, year) %>%
  summarise(mean_happiness = mean(Life.Ladder)) %>%
  ggplot(aes(x=year, y = mean_happiness, color = continent))  + geom_point() + geom_line() + scale_color_manual(values = cb_palette)+labs(y = "Average Happiness Index", x = "Year",
       title = "Chnage of Happiness over the years")
```


## Individual relationship of each independent variable with dependent variable

## Happiness Index vs Unemployment Rate
```{r}
Final_df %>%
  #summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate)) %>%
  ggplot(aes(x = Unemployment_Rate, y = Life.Ladder)) + geom_point() + geom_smooth(method = 'lm') + labs(y = "Average Happiness Index", x = "Avg Unemployment rate",
       title = "Avg Happiness index against Unemployment Rate")
```

## Happiness Index vs Life expectancy

```{r}
Final_df %>%
  #summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate)) %>%
  ggplot(aes(x = Life_expextancy, y = Life.Ladder)) + geom_point() + geom_smooth(method = 'lm') + labs(y = "Average Happiness Index", x = "Avg Life expectaancy",
       title = "Avg Happiness index against Avg Life expectancy")
```
## Happiness Index vs GDP percapita

```{r}
Final_df %>%
  #summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate)) %>%
  ggplot(aes(x = GDP_percapita, y = Life.Ladder)) + geom_point() + geom_smooth(method = 'lm') + labs(y = "Average Happiness Index", x = "Avg GDP percapita",
       title = "Avg Happiness index against Avg GDP percapita")
```


## Splitting data to before covid and during covid

```{r}
Final_df_2010_2018 = filter(Final_df, year <= '2018')
Final_df_2018_2021 = filter(Final_df, year > '2018')
```

## Trend of the Unemployment rate on Happiness Index grouped by continents
```{r}
Final_df %>%
  group_by(continent, Country) %>%
  #filter(continent != 'Ocenia') %>%
  summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate)) %>%
  ggplot(aes(x = mean_unemployment, y = mean_happiness, color = continent)) + geom_point() + geom_smooth(method = 'lm', se = FALSE) + scale_color_manual(values = cb_palette) + labs(y = "Average Happiness Index", x = "Average Unemployment Rate",
       title = "Average Happiness index against Unemployment Rate grouped by Continents")
```
## Trend of the Life expectancy on Happiness Index grouped by continents

```{r}
Final_df %>%
  group_by(continent, Country) %>%
  summarise(mean_happiness = mean(Life.Ladder), mean_lifeexpectancy = mean(Life_expextancy)) %>%
  ggplot(aes(x = mean_lifeexpectancy, y = mean_happiness, color = continent)) + geom_point() + geom_smooth(method = 'lm', se = FALSE) +  scale_color_manual(values = cb_palette) + labs(y = "Average Happiness Index", x = "Average Life expectancy",
       title = "Average Happiness index against Average Life expectancy grouped by Continents")
```
## Trend of the GDP percapita on Happiness Index grouped by continents

```{r}
Final_df %>%
  group_by(continent, Country) %>%
  summarise(mean_happiness = mean(Life.Ladder), mean_gdppercapita = mean(GDP_percapita)) %>%
  ggplot(aes(x = mean_gdppercapita, y = mean_happiness, color = continent)) + geom_point() + geom_smooth(method = 'lm', se = FALSE) + scale_color_manual(values = cb_palette)+ labs(y = "Happiness Index", x = "GDP percapita",
       title = "Happines vs GDP")
```


```{r}
lm(Life.Ladder ~ Unemployment_Rate, data  = Final_df_2010_2018)
lm(Life.Ladder ~ Unemployment_Rate, data  = Final_df_2018_2021)
```

```{r}
lm(Life.Ladder ~ Life_expextancy, data  = Final_df_2010_2018)
lm(Life.Ladder ~ Life_expextancy, data  = Final_df_2018_2021)
```


```{r}
lm(Life.Ladder ~ GDP_percapita, data  = Final_df_2010_2018)
lm(Life.Ladder ~ GDP_percapita, data  = Final_df_2018_2021)
```


## Difference in trend of unemployment rate over Happiness Index before and during covid
```{r}
temp = Final_df_2010_2018%>%
  group_by(Country) %>%
summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate)) %>%
  mutate(group.year = "2010-2018") 

temp_1 = Final_df_2018_2021%>%
  group_by(Country) %>%
summarise(mean_happiness = mean(Life.Ladder), mean_unemployment = mean(Unemployment_Rate)) %>%
  mutate(group.year = "2019-2021")

full_temp = union_all(temp, temp_1)

ggplot(full_temp, aes(x = mean_unemployment, y = mean_happiness, color = group.year)) + geom_point()  + geom_smooth(method="lm", span =0.8, se = FALSE) + scale_color_manual(values = cb_palette) + labs(y = "Average Happiness Index", x = "Average Unemployment Rate",
       title = "Avg Happiness index against Avg Unemployment Rate before and during covid")
```


## Difference in trend of Life expectancy over Happiness Index before and during covid

```{r}
temp = Final_df_2010_2018 %>%
  group_by(Country) %>%
summarise(mean_happiness = mean(Life.Ladder), mean_lifeexpectancy = mean(Life_expextancy)) %>%
  mutate(group.year = "2010-2018") 

temp_1 = Final_df_2018_2021 %>%
  group_by(Country) %>%
summarise(mean_happiness = mean(Life.Ladder), mean_lifeexpectancy = mean(Life_expextancy)) %>%
  mutate(group.year = "2019-2021")


full_temp = union_all(temp, temp_1)

ggplot(full_temp, aes(x = mean_lifeexpectancy, y = mean_happiness, color = group.year)) + geom_point()  + geom_smooth(method="lm", span =0.8, se = FALSE) + scale_color_manual(values = cb_palette) + labs(y = "Average Happiness Index", x = "Average Life expectancy",
       title = "Avg Happiness index against Avg Life expectancy before and during covid")
```



## Difference in trend of GDP percapita over Happiness Index before and during covid
```{r}
temp = Final_df_2010_2018 %>%
  group_by(Country) %>%
summarise(mean_happiness = mean(Life.Ladder), mean_gdppercapita = mean(GDP_percapita)) %>%
  mutate(group.year = "2010-2018")

temp_1 = Final_df_2018_2021 %>%
  group_by(Country) %>%
summarise(mean_happiness = mean(Life.Ladder), mean_gdppercapita = mean(GDP_percapita)) %>%
  mutate(group.year = "2019-2021") 

full_temp = union_all(temp, temp_1)

ggplot(full_temp, aes(x = mean_gdppercapita, y = mean_happiness, color = group.year)) + geom_point()  + geom_smooth(method="lm", span =0.8, se = FALSE) + scale_color_manual(values = cb_palette) + labs(y = "Average Happiness Index", x = "Average GDP percapita", title = "Avg Happiness index against Avg GDP percapita before and during covid")

```
## Predicting Happiness Index of Happiness Index during covid by training lm model on data during covid

```{r}
library(mgcv)
happiness_lm = lm(Life.Ladder ~ Life_expextancy + GDP_percapita + Unemployment_Rate, data = Final_df_2010_2018, familiy = 'symmetric')
plot(happiness_lm)
```


```{r}
happiness_grid = Final_df_2018_2021[,c("Life_expextancy", "GDP_percapita", "Unemployment_Rate")]
happiness.predict = predict(happiness_lm, newdata = happiness_grid)
```

```{r}
Final_df_2018_2021$predicted.happiness_lm = happiness.predict
```

```{r}
Final_df_2018_2021 %>%
  group_by(Country) %>%
  summarise(mean_happiness.predicted = mean(predicted.happiness_lm), mean_happiness.original = mean(Life.Ladder)) %>%
  merge(df_continents, by = c("Country")) %>%
  ggplot( aes(x =  mean_happiness.predicted, y = mean_happiness.original, color = continent)) + geom_point()  + geom_abline(slope = 1, intercept = 0) +geom_smooth(method="lm", span =0.8, se = FALSE)   + scale_color_manual(values = cb_palette)+labs(x = "Predicted Happiness score",
       y = "Actual Happiness Score",
       title = "Prediction of Happiness (lm model)")

```

## Mean squared error for lm model

```{r}
mean((Final_df_2018_2021$Life.Ladder - Final_df_2018_2021$predicted.happiness_lm)^2)
```


## Predicting Happiness Index of Happiness Index during covid by training gam model on data during covid
```{r}
happy_gam = gam(Life.Ladder ~s(Life_expextancy) + s(GDP_percapita) + s(Unemployment_Rate), data = Final_df_2010_2018)
```

```{r}
grid_gam = Final_df_2018_2021[,c("Life_expextancy", "GDP_percapita", "Unemployment_Rate")]
gam.predict = predict(happy_gam, newdata = grid_gam)
```

```{r}
Final_df_2018_2021$predicted.happy_gam = gam.predict
```

```{r}
Final_df_2018_2021 %>%
  group_by(Country) %>%
  summarise(mean_happiness.predicted = mean(predicted.happy_gam), mean_happiness.original = mean(Life.Ladder) ) %>%
  merge(df_continents, by = c("Country")) %>%
  ggplot( aes(x =  mean_happiness.predicted, y = mean_happiness.original, color = continent)) + geom_point()  + geom_abline(slope = 1, intercept = 0) +  geom_smooth(method = 'lm',se=FALSE)   + scale_color_manual(values = cb_palette)+labs(x = "Predicted Happiness Score ",
       y = "Actual Happiness Score",
       title = "Prediction of Happiness (GAM model)")
```
## Mean squared error for lm model
```{r}
mean((Final_df_2018_2021$Life.Ladder - Final_df_2018_2021$predicted.happy_gam)^2)
```


## Forecasting the avergae happiness index of the world during covid (2019-2021) using holt predictions 
```{r}
timeseries_df = Final_df %>%
  group_by(year) %>%
  summarise(mean_happiness = mean(Life.Ladder)) %>%
  filter(year < 2019)
```

```{r}
timeseries_object = ts(timeseries_df[,2], start = (2010), end = (2018), frequency = 1)
timeseries_object
```

```{r}
library(forecast)
library(fpp2)
library(TTR)
library(dplyr)
holt_pred = holt(timeseries_object, h = 3)
summary(holt_pred)
```
```{r}
holt_pred_df = as.data.frame(holt_pred)
```

```{r}
year = c(2019, 2020, 2021)
mean_happiness = holt_pred_df$`Point Forecast`
timeseries_df1 = data.frame(year, mean_happiness)
```


```{r}
timeseries_df = union_all(timeseries_df, timeseries_df1)
```

```{r}
timeseries_df  = timeseries_df  %>%
  rename(happiness.forecasted = mean_happiness)
```

```{r}
tempdf = Final_df %>%
  group_by(year) %>%
  summarise(mean_happiness = mean(Life.Ladder))
timeseries_df$original.happiness = tempdf$mean_happiness

```

```{r}
timeseries_df = timeseries_df %>%
  pivot_longer(cols = c(happiness.forecasted, original.happiness),
               names_to = 'happiness.type',
               values_to = 'happinness.value')
```



```{r}
ggplot(timeseries_df, aes(x = year, y = happinness.value , color = happiness.type)) + geom_point() + geom_line() + scale_color_manual(values = cb_palette) + labs(x = "Year",
       y = "Happiness Index",
       title = "Holts forecasting")
```


